Oregon: Zip-Level Wildfire Smoke and Health Outcomes (open in wildfire, not script)

### 2017-04-25 r_scripts/analysis/or_result_zip_manuscript.Rmd

This document contains results for the association between wildfire smoke PM~2.5~ and cardiovascular (CVD) and respiratory health outcomes. This particular markdown document focuses on the zip code-level analysis. Note there is a county-level analysis that will be paired with CDC mortality data as well. Herein, I compare different PM~2.5~ estimation methods described in further detail below.

Note, epidemiological methods that Sheryl and I think can improve our estimation and reduce some bias around the point estimates will not be the focus of the manuscript.

```{r data import, include = F, echo = FALSE}
# Set working directory and read in files --------------------------------------
# relative path
# set working directory if you want to run seperate chunks
# setwd("./r_scripts")
getwd()

path <- paste0("../data_new/health_data")
# Infile case-crossover dataframes ---------------------------------------------
# Dataframes made in 'chars_3month_binary_smoke_may2016 script
# resp exacerbations
resp_casecross <- read_csv(paste(path, 
                        "respiratory_may_to_sep_time_strat_casecross.csv", 
                        sep = "/"))

# asthma
asthma_casecross <- read_csv(paste(path, 
                          "asthma_may_to_sep_time_strat_casecross.csv", 
                          sep = "/"))

# copd 
copd_casecross <- read_csv(paste(path, 
                        "copd_may_to_sep_time_strat_casecross.csv", 
                        sep = "/"))

# pneum or bronchitis
pneum_casecross <- read_csv(paste(path, 
                        "copd_may_to_sep_time_strat_casecross.csv",
                        sep="/"))
# acute bronchitis
acute_bronch_casecross <- read_csv(paste(path, 
                                "acute_bronch_may_to_sep_time_strat_casecross.csv",
                                sep = "/"))
# cvd
cvd_casecross <- read_csv(paste(path, 
                      "cvd_may_to_sep_time_strat_casecross.csv",
                      sep="/"))
# arrhythmia
arrhythmia_casecross <- read_csv(paste(path, 
                              "arrhythmia_may_to_sep_time_strat_casecross.csv",
                               sep="/"))
# cerebral vascular
cereb_vas_casecross <- read_csv(paste(path, 
                              "cerbrovas_dis_may_to_sep_time_strat_casecross.csv",
                              sep="/"))
# heart failure
hf_casecross <- read_csv(paste(path, 
                      "heart_failure_may_to_sep_time_strat_casecross.csv", 
                      sep="/"))
# ischemic heart disease
ihd_casecross <- read_csv(paste(path, 
                       "isch_heart_dis_may_to_sep_time_strat_casecross.csv",
                       sep="/"))
# myo infarc
mi_casecross <- read_csv(paste(path, 
                      "myocardial_infarc_may_to_sep_time_strat_casecross.csv", sep="/"))

# broken arm
broken_arm_casecross <- read_csv(paste(path, 
                              "broken_arm_may_to_sep_time_strat_casecross.csv",
                              sep="/"))

# read zip pm for counts by zip
smoke_path <- paste0("../data/Oregon/")

zip_pm <- read_csv(paste0(smoke_path, "zip_pm_to_merge_with_chars.csv"))

```{r, eval = FALSE}
## 
# read locations of fires
fire_loc <- read_csv(paste0(smoke_path, "washington_fires_locations201209.csv"))

# import the zip-level met data files Rish sent
zip_met <- read_csv(paste0(smoke_path, "wa_zip_exp_0710_2012_ryan.csv")) %>% 
  # define date variable; original character looked like: 01JUL2012
  mutate(date = as.Date(date, "%d%B%Y"))

# note 12/4/16: read_csv is pretty slow on pc, fread doesn't import date variable 
# that well. Solution may be to use fread but convert to date to use later

@@

Note for following tables: I wrote out 'Greater than rather' than using  '>' as it was increasing the following font and looked weird.

@@

# set list and do cross tabs to find values
# Producing conditional logit model estimates loop 
for(i in 1:length(df_list)){

  i = 2
# dataframe to loop through
  df_to_loop <- data.frame(df_list[i])
  # indication of column
  outcome <- colnames(df_to_loop[76])
  # outcome name
  outcome_name <- outcome_list[i]

@@

tab <- htmlTable(txtRound(combined_point_est_df, digits = 1), 
           caption = "Number of cases for each outcome observed from July 1st to October 31st, 2012",
           # column headers
           header = c("Outcome", "Cases n", "Less than 15", "(%)", "15 to 65", "(%)", "Greater than 65",
                      "(%)", "Female", "(%)", "Male", "(%)"),
           # column spanner
           cgroup = c("","Age Category", "Sex"), 
           n.cgroup = c(2, 6, 4),
           padding.rgroup = "&nbsp;&nbsp;",
           css.cell = "padding-left: 0.5em; padding-right: .5em;", # cell space
           align = "llccccc" # column alignment,
            ) # end table

@@

### Background on Time-Stratified Case-Crossover Design

The original time-stratified case-crossover design suggested picking referent periods based on any potential time-varying confounding, *a priori*. For example, if day of the week and the month of the exposure sequence is of concern, one might pick referents in the same day of the week and the same day of the month. The papers that discuss and support the use of a time-stratified design are the Janes et al. 2005 **Epidemiology paper**, *Case-Crossover Analyses of Air Pollution Exposure Data; Referent Selection Strategies and Their Implication for Bias*.

The authors state that with the time-stratified design, the *a priori* selection of the referent period on the same day of the week within the same month has the following benefits:

1. Localizable - Meaning that the likelihood of index times conditional on the referent windows contains information about \beta, which I think means the relative difference in the exposed vs unexposed, conditioned on the exposure (air pollution in their example), is unbiased.

2. Non-ignroable - Meaning that time-varying factors can be ignored when using a conditional logistic regression; controlled through design (matching on day of week within the same season).

More referent periods is better for statistical efficiency, but there is a tradeoff, where additional referent periods may introduce what they call 'overlap bias', which they mathmatically define in Janes 2005 **Statistics in Medicine** paper, *Overlap Bias in the Case-Crossover Design, with Applications to Air Pollution Exposures*. Basically this bias occurs when the exposure assigned in the referent period is conditional on the strategy used to define the referent period.

As Janes et al. state in their 2005 **Statistics in Medicine** paper:

  > *Two of the main strengths of the case-crossover design in the air pollution context are
that it controls fixed confounders and allows for control over time-dependent confounders
by design. Choosing referents that are restricted to the same day of the week and season
as the index time will control for these effects. Though we have not examined the degree
to which various referent selection strategies control bias due to confounding, several points
are clear. Referents that are closer to the index time will achieve greater control of seasonal
confounding. However, sampling referents closer to the index time will also result in a loss
of power due to less heterogeneity in the exposure series, and a smaller number of referents
will also result in lower power.*

I will use both Janes 2005 papers in **Epidemiology** and **Statistics in Medicine** to justify the use of the time-stratified design.

This design has an advantage over other case-crossover designs I've explored in that you can use more referent times, and as the exposure at these times are averaged for each subject over a set time period, this balances out some of the more extreme values of PM~2.5~ that may persist over a couple weeks due to wildfire smoke.

### Time-Stratified Case-Crossover (referents same day of week within fire season July - October)

For these analyses and for the final results in the paper, I used referents within the entire wildfire season (July to October). This time-stratified referent strategy allows even more referent observations to average to 'smooth' out some of those extreme PM~2.5~ concentrations in some referent periods. For my sensitivity analysis of varying referent periods, please see other scripts.

Adjusting for CDC Met Variables: Rish has provided zip-level meteorologic data on wind speed, precipitation, relative humidity, and temperature. Note a slight difference that will likely need to be addressed in discussion is that these measures are weighted based on census tract, not on the WRF grid like our measures of smoke.

```{r data wrangle with zip met vars, warning = F, echo = F, message = F}

# create dataset that can be merged to case-cross over dataframes
daily_zip_met <- zip_met %>%
  # limit to new mean met variables (getting rid of smoke variables I already have)
  select(1:7) %>% 
  rename(ZIPCODE = ZIP) %>% arrange(ZIPCODE, date) %>%
  group_by(ZIPCODE) %>% 
  # create lag variables
  mutate(
    # absolute temp
    meanAT_lag1 = lag(daily_meanAT, 1, order_by = ZIPCODE), 
    meanAT_lag2 = lag(daily_meanAT, 2, order_by = ZIPCODE),
    meanAT_lag3 = lag(daily_meanAT, 3, order_by = ZIPCODE),
    meanAT_lag4 = lag(daily_meanAT, 4, order_by = ZIPCODE),
    meanAT_lag5 = lag(daily_meanAT, 5, order_by = ZIPCODE),
    # relative humidity
    meanRH_lag1 = lag(daily_meanRH, 1, order_by = ZIPCODE),
    meanRH_lag2 = lag(daily_meanRH, 2, order_by = ZIPCODE),
    meanRH_lag3 = lag(daily_meanRH, 3, order_by = ZIPCODE),
    meanRH_lag4 = lag(daily_meanRH, 4, order_by = ZIPCODE),
    meanRH_lag5 = lag(daily_meanRH, 5, order_by = ZIPCODE),
    # temp
    meanT_lag1 = lag(daily_meanT, 1, order_by = ZIPCODE),
    meanT_lag2 = lag(daily_meanT, 2, order_by = ZIPCODE),
    meanT_lag3 = lag(daily_meanT, 3, order_by = ZIPCODE),
    meanT_lag4 = lag(daily_meanT, 4, order_by = ZIPCODE),
    meanT_lag5 = lag(daily_meanT, 5, order_by = ZIPCODE),
    # precip
    meanPrcp_lag1 = lag(daily_meanPrcp, 1, order_by = ZIPCODE),
    meanPrcp_lag2 = lag(daily_meanPrcp, 2, order_by = ZIPCODE),
    meanPrcp_lag3 = lag(daily_meanPrcp, 3, order_by = ZIPCODE),
    meanPrcp_lag4 = lag(daily_meanPrcp, 4, order_by = ZIPCODE),
    meanPrcp_lag5 = lag(daily_meanPrcp, 5, order_by = ZIPCODE),
    # wind speed
    meanWS_lag1 = lag(daily_meanWS, 1, order_by = ZIPCODE),
    meanWS_lag2 = lag(daily_meanWS, 2, order_by = ZIPCODE),
    meanWS_lag3 = lag(daily_meanWS, 3, order_by = ZIPCODE),
    meanWS_lag4 = lag(daily_meanWS, 4, order_by = ZIPCODE),
    meanWS_lag5 = lag(daily_meanWS, 5, order_by = ZIPCODE)) %>% 
  # sort so all like variables are together
  select(1:3, 8:12, 4, 13:17, 5, 18:22, 6, 23:27, 7, 28:32)

#which(colnames(daily_zip_met)=="meanAT_lag5")
#glimpse(daily_zip_met)

```{r time strat within fire season adj for cdc met, warning = F, echo = F, results='asis'} 
# note above, results = 'asis' needed to print htmlTables

# I removed copd exacerbation and rheuamtoid arthritis

@@

  print(print_plot)
  # save figure
  ggsave("./fig4.pdf", plot = print_plot, 
       width = 12, height = 8, units = "in")

# figures for research day
  # plot of only respiratory outcomes

  xtabs(~ combined_point_est_df$outcome)
# subset out global method and respiratory outcomes ----
resp_point_est_df <- combined_point_est_df %>%

  filter(outcome == "All Respiratory" |
         outcome == "Asthma" |
         outcome == "COPD" |
         outcome == "Pneumonia" |
         outcome == "Acute Bronchitis" |
         outcome == "Broken Arm")

## ggplot
  resp_plot <- ggplot(resp_point_est_df, 
                      aes(x = pm_method, y = odds_ratio, colour = pm_method)) +
    # custom color 
    scale_color_manual(name = "Smoke-Estimation Method", 
                       values = c("red", "blue", "#32115C"),
                       guide = guide_legend(title.position = "top", title.hjust = 0.5)) +
    geom_point() + #geom_text(vjust = 0, nudge_x = 0.3) +
    geom_errorbar(aes(ymin=lower95, ymax=upper95), width = 0.2) +
    facet_wrap(~outcome, nrow = 2) +
    geom_hline(yintercept = 1, linetype=2) +
#    ggtitle('Association Between PM2.5 \n from Wildfire Smoke on Hospitalizations') +
    ylab(expression(paste("Odds Ratio for 10µg/m"^3, " Increase in PM"[2.5]))) +
    #xlab('Smoke Estimation Method') +
    # plot theme
    theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    # strip element
    strip.background = element_rect(colour=NA, fill=NA),
    panel.border = element_rect(fill = NA, color = "black"),
    # facet text size
    #strip.text = element_text(size = 8),
    # axis element
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.y = element_text(angle = 90),
    # legend elements
    legend.position = "bottom")
    #legend.text = element_text(size = 8))

  print(resp_plot)

# subset out global method and respiratory outcomes ----
cvd_point_est_df <- combined_point_est_df %>% 
   filter(outcome == "Cardiovascular Disease" |
         outcome == "Arrhythmia" |
         outcome == "Cerebrovascular Disease" |
         outcome == "Heart Failure" |
         outcome == "Ischemic Heart Disease" |
         outcome == "Myocardial Infarction")

## ggplot
  cvd_plot <- ggplot(cvd_point_est_df, 
                      aes(x = pm_method, y = odds_ratio, colour = pm_method)) +
    # custom color 
    scale_color_manual(name = "Smoke-Estimation Method", 
                       values = c("red", "blue", "#32115C"),
                       guide = guide_legend(title.position = "top", title.hjust = 0.5)) +
    geom_point() + #geom_text(vjust = 0, nudge_x = 0.3) +
    geom_errorbar(aes(ymin=lower95, ymax=upper95), width = 0.2) +
    facet_wrap(~outcome, nrow = 2) +
    geom_hline(yintercept = 1, linetype=2) +
#    ggtitle('Association Between PM2.5 \n from Wildfire Smoke on Hospitalizations') +
    ylab(expression(paste("Odds Ratio for 10µg/m"^3, " Increase in PM"[2.5]))) +
    #xlab('Smoke Estimation Method') +
    # plot theme
    theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    # strip element
    strip.background = element_rect(colour=NA, fill=NA),
    panel.border = element_rect(fill = NA, color = "black"),
    # facet text size
    #strip.text = element_text(size = 8),
    # axis element
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.y = element_text(angle = 90),
    # legend elements
    legend.position = "bottom")
    #legend.text = element_text(size = 8))

  print(cvd_plot)

#### Result Summary: Time-Stratified Case-Crossover Design (same day of week within the season)

A general interpretation is as follows:
A 10 $\mu$g/m^3^ increase in PM~2.5~ attributed to wildfire smoke was associated with a X% increase in the risk for a health event. Where X% is the odds ratio - 1 (e.g. OR of 1.20 - 1 = 20% increase). 
Note that the odds ratio approximates the relative risk in this case due to the rare outcome assumption in that the calcuation for an odds ratio is similar to risk ratio calcuation.

We see that all exposure methods are significantly associated with an increase in all respiratory outcomes and asthma. The WRF-Chem smoke method is associated with cerebrovascular disease as well. As for the other models, Kriging, Global Regression, and Geo-Weighted Regression, those are associated with COPD. I believe this is the most justifiable design approach that would average any extreme values of PM~2.5~ across the wildfire season for the vector of referent values for a particular subject.

When adjusting for the CDC met variables, the results are slightly attenuated when compared to my adjustments with just WRF-Temp, but my overall conclusions do not change much.

Here is the same table, but within season instead of fire season and adjusting for CDC met variables. Rish is worried about time-variant confounding within a subject that may bias the overall results. This might include seasonal differences in baseline risk of asthma.

@@

# looking only at wrf chem, kriging, and geo-weighted; taking out 'Global Smoke' method
method_list <- c('WRF-Chem Smoke', 'Kriging Smoke', 'Global Smoke', 'GWR Smoke')

@@

@@

# dataframe to loop through
  df_to_loop <- data.frame(df_list[i])
  # indication of column
  outcome <- colnames(df_to_loop[3])
  # outcome name
  outcome_name <- outcome_list[i]

  # extract covariates from dataframe
  covariates_df <- df_to_loop[, c(1:16, 27:28, 151:157)]

  # extract pm values and divide by 10 and ordered
  #which(colnames(df_to_loop)=="global_smk_pm_zip") # code to find column numbers
  pm_estimates_df <- df_to_loop[, c(19, 26, 25, 24)]/10  # create 10 unit increases

@@

    # empty matrix (12 x 10 matrix)
    point_estimates <- matrix(nrow = 4, ncol = 10, byrow = T)

@@

    # dataframe for analysis creation
    # bind columns back together 
    df_analysis <- cbind(covariates_df, pm_estimates_df) %>% 
      # remove missing pm values
      filter(!is.na(wrf_smk_pm_zip)) %>% 
      # join in cdc met data
      full_join(daily_zip_met, by = c("ZIPCODE", "date")) %>% 
      # limit to emergency or urgent care
      filter(ADM_TYPE == 1 | ADM_TYPE == 2) %>%
      # limit to specific age category
      filter(age_cat == k) %>% 
      # the following code makes sure that the counterfactual values retained are 
      # symetric in that number of obs before = number of obs after
      mutate(obs_diff_admission = (date - date_admit)/7) 
      # dataframe is already for the entire fire season, so I don't need to subset anymore

  # second loop to run a model for each pm estimation method
    for(j in 26:29){

@@

      # set row number to fill
      row_n <- j-25

      # only run the model if the dataframe has observations
      if(nrow(df_analysis) != 0){
      # conditional logistic regression model
      mod <- clogit(outcome ~ df_analysis[[j]] + daily_meanRH + daily_meanT + daily_meanWS + 
                    daily_meanPrcp + strata(PATIENTID), df_analysis)

@@

  print(print_plot)
  # save figure
  ggsave("./fig5.pdf", plot = print_plot, 
       width = 12, height = 8, units = "in")

@@

    # dataframe for analysis creation
    # bind columns back together 
    df_analysis <- cbind(covariates_df, pm_estimates_df) %>% 
      # remove missing pm values
      filter(!is.na(wrf_smk_pm_zip)) %>% 
      # join in cdc met data
      full_join(daily_zip_met, by = c("ZIPCODE", "date")) %>% 
      # limit to emergency or urgent care
      filter(ADM_TYPE == 1 | ADM_TYPE == 2) %>%
      # limit to specific sex category
      filter(sex_num == k) %>% 
      # the following code makes sure that the counterfactual values retained are 
      # symetric in that number of obs before = number of obs after
      mutate(obs_diff_admission = (date - date_admit)/7) 
      # dataframe is already for the entire fire season, so I don't need to subset anymore

  # second loop to run a model for each pm estimation method
    for(j in 26:29){

@@

  print(print_plot)
  # save figure
  ggsave("./fig6.pdf", plot = print_plot, 
       width = 12, height = 8, units = "in")

@@

### Lag Periods

Lag period adjusted for CDC met variables.

```{r lags with cdc met adj, warning = F, echo = F, results='asis'} 
# note above, results = 'asis' needed to print htmlTables

# looking only at wrf chem, kriging, and geo-weighted; taking out 'Global Smoke' method
method_list <- c('WRF-Chem Smoke', 'Kriging Smoke', 'Global Smoke', 'GWR Smoke')

# dataframe to loop through
  df_to_loop <- data.frame(df_list[i])
  # indication of column
  outcome <- colnames(df_to_loop[3])
  # outcome name
  outcome_name <- outcome_list[i]
  # extract covariates from dataframe
  covariates_df <- df_to_loop[, c(1:16, 27,79:83, 28, 151:157)]

  # extract pm values and divide by 10 and ordered
  #which(colnames(df_to_loop)=="wrf_temp_lag1_zip") # code to find column numbers
  #which(colnames(df_to_loop)=="wrf_temp_zip") # code to find column numbers
  pm_lag_list <- list(df_to_loop[, c(19, 39:43)]/10, df_to_loop[, c(26, 74:78)]/10,
                  df_to_loop[, c(25, 69:73)]/10, df_to_loop[, c(24, 64:68)]/10)  # create 10 unit increases

# need to think if I need a methods section (probably not since this is it)

  # new loop for each element of the lag list 
  for(k in 1:4){

    # empty matrix (12 x 10 matrix)
    point_estimates <- matrix(nrow = 6, ncol = 10, byrow = T)

    colnames(point_estimates) <- c('outcome', 'pm_method', 'lag', 'n', 'n_events', 
                                   'odds_ratio', 'lower95', 'upper95', 'se', 'p_val')

    # fill in the outcome namedataframe before method loop
    point_estimates[, 1] <- outcome_name
    # repeat the sex category 4 times for each pm method
    point_estimates[, 2] <- method_list[k]

    pm_test <- pm_lag_list[[k]]

    # dataframe for analysis creation
    # bind columns back together 
    df_analysis <- cbind(covariates_df, pm_lag_list[[k]]) %>% 
      # limit to emergency or urgent care
      filter(ADM_TYPE == 1 | ADM_TYPE == 2) %>%
      # join in cdc met data
      full_join(daily_zip_met, by = c("ZIPCODE", "date")) 
      # dataframe is already for the entire fire season, so I don't need to subset anymore
#glimpse(df_analysis)

  # loop to run a model for each lag 0-5
    for(j in 0:5){

      # finding locations of columns in dataframe
      which(colnames(df_analysis)=="daily_meanWS")

      # set column number to evaluate for the pm lag method
      pm_lag_colnum <- j + 31
      # set column number of cdc temp lag to adjust for
      temp_lag_colnum <- j + 49
      # set col num for abs temp
      ab_temp_lag_colnum <- j + 37
      # set col num for relative humidity
      rh_lag_colnum <- j + 43
      # precip
      pr_lag_colnum <- j + 55
      # wind speed
      ws_lag_colnum <- j + 61
      # set row number to fill
      row_n <- j+1

      # only run the model if the dataframe has observations
      if(nrow(df_analysis) != 0){
      # conditional logistic regression model
      mod <- clogit(outcome ~ df_analysis[[pm_lag_colnum]] + df_analysis[[temp_lag_colnum]] + 
                    df_analysis[[rh_lag_colnum]] + df_analysis[[pr_lag_colnum]] + 
                    df_analysis[[ws_lag_colnum]] + strata(PATIENTID), df_analysis)

      # populate matrix
      point_estimates[row_n, 3] <- j
      point_estimates[row_n, 4] <- mod$n
      point_estimates[row_n, 5] <- mod$nevent
      # odds ratio
      point_estimates[row_n, 6] <- round(exp(summary(mod)$coefficient[1,1]), 3)
      # 95% lower bound
      point_estimates[row_n, 7] <- round(exp((summary(mod)$coefficient[1,1]) -
                                        1.96*(summary(mod)$coefficient[1,3])), 3)
      # 95% upper bound
      point_estimates[row_n, 8] <- round(exp((summary(mod)$coefficient[1,1]) +
                                        1.96*(summary(mod)$coefficient[1,3])), 3)
      # standard error
      point_estimates[row_n, 9] <- round(summary(mod)$coefficient[1,3], 4)
      # p val
      point_estimates[row_n, 10] <- round(summary(mod)$coefficient[1,5], 4)

  # bind rows of sex category estimates together
 lag_est_df <- bind_rows(datalist1)

 # populate second dataframe list
 datalist2[[i]] <- lag_est_df

#glimpse(combined_point_est_df)
# subset to just geo-smoke estimates for now to reduce number of rows
wrf_geo_lag <- combined_point_est_df %>% 
  # filter columns to just geo-smk
  filter(pm_method == "GWR Smoke" |
         pm_method == "WRF-Chem Smoke") %>% 
  # subset columns I want to put in to the table
  select(2:3, 5:8) %>% 
  mutate(odds_ratio = ifelse(odds_ratio == 99, "--", odds_ratio),
         lower95 = ifelse(lower95 == 99, "--", lower95),
         upper95 = ifelse(upper95 == 99, "--", upper95))

  tab <- htmlTable(txtRound(wrf_geo_lag, digits = 3, 1:3), 
           caption = "Association between a 10 ug/m^3 in PM2.5 geo smoke and health outcomes by lagged days",
           # row group by outcome
           rgroup = outcome_list,
           n.rgroup = c(rep(12, 12)), # 6 rows for each lag period for each outcome
           # column headers
           header = c("Est Method", "Lagged Days", "Events",
                      "OR&dagger;", "Lower", "Upper"),
           # column spanner
           cgroup = c("", "95% CI"), 
           n.cgroup = c(4, 2),
           padding.rgroup = "&nbsp;&nbsp;",
           css.cell = "padding-left: 0.5em; padding-right: .5em;", # cell space
           align = "lccccc", # column alignment,
           tfoot="&dagger; Adjusted for lagged CDC temperature, windspeed, relative humidity, and precipitation, accounting for subject."
            ) # end table

# ggplot of odds ratios, facet wrapped by outcomes -----
# convert variables from character to either numeric or factor
# factor preserves the order of the variable
wrf_geo_lag_plot <- combined_point_est_df %>% 
  # filter columns to just geo-smk
  filter(pm_method == "GWR Smoke" |
         pm_method == "WRF-Chem Smoke") %>% 
  # subset columns I want to put in to the table
  select(1:3, 5:8)

# change characters to numeric and factor  
wrf_geo_lag_plot$outcome <- factor(wrf_geo_lag_plot$outcome,
                              levels = unique(wrf_geo_lag_plot$outcome))

wrf_geo_lag_plot$pm_method <- factor(wrf_geo_lag_plot$pm_method,
                                levels = unique(wrf_geo_lag_plot$pm_method))

wrf_geo_lag_plot$lag <- factor(wrf_geo_lag_plot$lag,
                                levels = unique(wrf_geo_lag_plot$lag))

wrf_geo_lag_plot$n_events <- as.numeric(wrf_geo_lag_plot$n_events)
wrf_geo_lag_plot$odds_ratio <- as.numeric(wrf_geo_lag_plot$odds_ratio)
wrf_geo_lag_plot$lower95 <- as.numeric(wrf_geo_lag_plot$lower95)
wrf_geo_lag_plot$upper95 <- as.numeric(wrf_geo_lag_plot$upper95)

## ggplot
  print_plot <- ggplot(wrf_geo_lag_plot,
                  aes(x = lag, y = odds_ratio, color = pm_method), na.rm = T) +
    geom_point(position = position_dodge(width = 0.5), na.rm = T) + 
    geom_errorbar(aes(ymin=lower95, ymax=upper95), 
                  position = position_dodge(width = 0.5), width = 0.2, na.rm =T) +
    # custom color 
    scale_color_manual(name = "Smoke-Estimation Method", 
                       values = c("red", "#32115C"),
                       guide = guide_legend(title.position = "top", title.hjust = 0.5)) +
    facet_wrap(~outcome, nrow = 3, scales = "free") +
    geom_hline(yintercept = 1, linetype=2) +
    #ggtitle('Association Between PM2.5 from \n Wildfire Smoke on Hospitalizations') +
    ylab(expression(paste("Odds Ratio for 10µg/m"^3, " Increase in PM"[2.5]))) +
    #ylim(0.7, 1.4) +
    xlab('Lag Days') +
    # plot theme
    theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    # strip element
    strip.background = element_rect(colour=NA, fill=NA),
    panel.border = element_rect(fill = NA, color = "black"),
    # facet text size
    strip.text = element_text(size = 10),
    # axis element
    #axis.text.x = element_blank(),
    #axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.y = element_text(angle = 90),
    # legend elements
    legend.position = "bottom")
    #legend.text = element_text(size = 8))

  print(print_plot)

>>>>

This document contains results for the association between wildfire smoke PM~2.5~ and cardiovascular (CVD) and respiratory health outcomes. This particular markdown document focuses on the zip code-level analysis. Herein, I compare different PM~2.5~ estimation methods described in further detail below.

### Time-Stratified Case-Crossover (referents same day of week within fire season May - September)

```{r include=FALSE}
library(tidyverse)
library(data.table)
library(survival)
library(htmlTable)
library(readr)
```

```{r data import, include = F, eval = FALSE}

### Limit to only ER visits
## pos = 23
var_list <- c('respiratory', 'asthma', 'pneumonia',  'acute_bronch', 'copd', 
              'cvd', 'isch_heart_dis', 'arrhythmia', 'heart_failure', 
              'cerbrovas_dis', 'myocardial_infarc', 'broken_arm')

N <- NA
n <- 1

for (i in var_list){
  read_path <- paste('../data_new/update/ER/or', i, 'may_to_sep_time_strat_casecross.csv', sep = "_")
  or_disease <- read_csv(read_path)

  or_disease <- or_disease %>% 
    filter(pos=="23") %>%
    # filter the unknown gender
    filter(gender!= "U")

  # Create a permanent case-cross over dataset
  file_name <- paste('../data_new/update/ER/or', i, 'may_to_sep_time_strat_casecross_er.csv', sep = "_")

  # write permanent dataset
  write_csv(or_disease, paste0(file_name))
  N[n] = length(unique(or_disease$personkey))
  n = n+1
}

num_person <- data.frame(rbind(var_list, N))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
## Analysis
path <- paste0("../data_new/update/ER")

resp_casecross <- read_csv(paste(path, 
                                 "or_respiratory_may_to_sep_time_strat_casecross_er.csv", 
                                 sep = "/"))

# asthma
asthma_casecross <- read_csv(paste(path, 
                                   "or_asthma_may_to_sep_time_strat_casecross_er.csv", 
                                   sep = "/"))

# copd 
copd_casecross <- read_csv(paste(path, 
                                 "or_copd_may_to_sep_time_strat_casecross_er.csv", 
                                 sep = "/"))

# pneum or bronchitis
pneum_casecross <- read_csv(paste(path, 
                                  "or_pneumonia_may_to_sep_time_strat_casecross_er.csv",
                                  sep="/"))
 # acute bronchitis
acute_bronch_casecross <- read_csv(paste(path, 
                                         "or_acute_bronch_may_to_sep_time_strat_casecross_er.csv",
                                         sep = "/"))
# cvd
cvd_casecross <- read_csv(paste(path, 
                                "or_cvd_may_to_sep_time_strat_casecross_er.csv",
                                sep="/"))
# arrhythmia
arrhythmia_casecross <- read_csv(paste(path, 
                                       "or_arrhythmia_may_to_sep_time_strat_casecross_er.csv",
                                       sep="/"))
# cerebral vascular
cereb_vas_casecross <- read_csv(paste(path, 
                                      "or_cerbrovas_dis_may_to_sep_time_strat_casecross_er.csv",
                                      sep="/"))
# heart failure
hf_casecross <- read_csv(paste(path, 
                               "or_heart_failure_may_to_sep_time_strat_casecross_er.csv", 
                               sep="/"))
# ischemic heart disease
ihd_casecross <- read_csv(paste(path, 
                                "or_isch_heart_dis_may_to_sep_time_strat_casecross_er.csv",
                                sep="/"))
# myo infarc
mi_casecross <- read_csv(paste(path, 
                               "or_myocardial_infarc_may_to_sep_time_strat_casecross_er.csv", sep="/"))

# broken arm
broken_arm_casecross <- read_csv(paste(path, 
                                       "or_broken_arm_may_to_sep_time_strat_casecross_er.csv",
                                       sep="/"))

# read zip pm for counts by zip
smoke_path <- paste0("../data/Oregon/")
zip_pm <- read_csv(paste0(smoke_path, "zip_pm_to_merge_with_chars.csv"))
```

@@

For all of 2013, there were a total of millons of all claims records in the APAC dataset. During the May 1st to September 30th wildfire season in Oregon state, there were 47413 emergency visits for the specified cardiopulmonary codes. The number of persons (cases), as well as age-specific and sex-specific strate.

@@

# set list and do cross tabs to find values
# Producing conditional logit model estimates loop 
for(i in 1:length(df_list)){
# dataframe to loop through
  df_to_loop <- data.frame(df_list[i])
  # indication of column
  outcome <- colnames(df_to_loop[76])
  # outcome name
  outcome_name <- outcome_list[i]

@@

tab <- htmlTable(txtRound(combined_point_est_df, digits = 1), 
           caption = "Number of cases for each outcome observed from May 1st to September 30st, 2013",
           # column headers
           header = c("Outcome", "Cases n", "Less than 15", "(%)", "15 to 65", "(%)", "Greater than 65",
                      "(%)", "Female", "(%)", "Male", "(%)"),
           # column spanner
           cgroup = c("","Age Category", "Sex"), 
           n.cgroup = c(2, 6, 4),
           padding.rgroup = "&nbsp;&nbsp;",
           css.cell = "padding-left: 0.5em; padding-right: .5em;", # cell space
           align = "llccccc" # column alignment,
            ) # end table

@@

```{r echo = F, warning = F, results='asis'}

df_list <- list(resp_casecross, asthma_casecross, copd_casecross, 
                pneum_casecross, acute_bronch_casecross, cvd_casecross, arrhythmia_casecross,
                cereb_vas_casecross, hf_casecross, ihd_casecross, mi_casecross, 
                broken_arm_casecross)

@@

print(print_plot)
# save figure
ggsave("./fig_only_er.pdf", plot = print_plot, 
       width = 12, height = 8, units = "in")
```

@@

# looking only at wrf chem, kriging, and geo-weighted; taking out 'Global Smoke' method
method_list <- c('WRF-Chem Smoke', 'Kriging Smoke', 'GWR Smoke')

@@

var_list <- c('respiratory', 'asthma', 'pneumonia',  'acute_bronch', 'copd', 
              'cvd', 'isch_heart_dis', 'arrhythmia', 'heart_failure', 
              'cerbrovas_dis', 'myocardial_infarc', 'broken_arm')

@@

# dataframe to loop through
  df_to_loop <- data.frame(df_list[i])
  # indication of column
  outcome <- var_list[i]
  # outcome name
  outcome_name <- outcome_list[i]

  # extract pm values and divide by 10 and ordered
  #which(colnames(df_to_loop)=="global_smk_pm_zip") # code to find column numbers
  pm_estimates_df <- df_to_loop[, c(88, 93, 92, 94)]/10  # create 10 unit increases

@@

    # empty matrix (12 x 10 matrix)
    point_estimates <- matrix(nrow = 3, ncol = 10, byrow = T)

@@

    # dataframe for analysis creation
    # bind columns back together 
    df_analysis <- cbind(covariates_df, pm_estimates_df) %>% 
      # remove missing pm values
      filter(!is.na(wrf_smk_pm_zip)) %>% 
      # limit to specific age category
      filter(age_ind == k) %>% 
      # the following code makes sure that the counterfactual values retained are 
      # symetric in that number of obs before = number of obs after
      mutate(obs_diff_admission = (fromdate - date)/7) 
      # dataframe is already for the entire fire season, so I don't need to subset anymore

@@

      # set row number to fill
      row_n <- j-38

      # only run the model if the dataframe has observations
      if(nrow(df_analysis) != 0){
      # conditional logistic regression model
      mod <- clogit(outcome ~ df_analysis[[j]] + wrf_temp_zip + strata(personkey), df_analysis)

@@

  print(print_plot)
  # save figure
  ggsave("./age_fig.pdf", plot = print_plot, 
       width = 12, height = 8, units = "in")

@@

    # dataframe for analysis creation
    # bind columns back together 
    df_analysis <- cbind(covariates_df, pm_estimates_df) %>% 
      # remove missing pm values
      filter(!is.na(wrf_smk_pm_zip)) %>% 
      # limit to specific sex category
      filter(sex_ind == k) %>% 
      # the following code makes sure that the counterfactual values retained are 
      # symetric in that number of obs before = number of obs after
      mutate(obs_diff_admission = (fromdate - date)/7) 
      # dataframe is already for the entire fire season, so I don't need to subset anymore

@@

  print(print_plot)
  # save figure
  ggsave("./sex_fig.pdf", plot = print_plot, 
       width = 12, height = 8, units = "in")

@@
